The Impact of the 2022-23 NBA Rule Change

### 2024-04-24 paper/paper.qmd
Update paper
Update discussion, abstract, and introduction sections

---
title: "The Impact of the 2022-23 NBA Rule Change"
subtitle: "Exploring the Relationship Between the New Transitional Foul Rule, Player Shooting Performance, and MVP Points"
author: 
  - Amie Liu
thanks: "Code and data are available at: https://github.com/AmieLiu/the-NBA-Rule-Change.git."
date: today
date-format: long
abstract: "Rule changes in professional basketball play a critical role in shaping both strategies and player performances. Specifically, the introduction of the transitional foul rule in the 2022-23 NBA season, which changes the way fouls are called during transition plays, could significantly affect players' shooting efficiency and their prospects for MVP awards. This paper examines the implications of this rule change, comparing player statistics between the 2022-23 and 2021-22 seasons. We find significant correlations between shooting efficiency and MVP points, highlighting the rule's influence on player recognition. This paper offers insights into the interplay between rule modifications, player performance, and individual accolades, underscoring the broader significance of adapting to regulatory shifts in professional sports."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

A clear gap exists in understanding the nuanced impacts of the new transitional foul rule on player performance and MVP considerations, especially when compared to the previous season. This paper seeks to fill this gap by thoroughly analyzing the statistical data from both the 2022-23 and 2021-22 NBA seasons to highlight the changes, continuities, and emerging trends over these times. To address this gap, we used the R programming language [@citeR] to analyze shooting performance metrics, such as field goal percentage (FG%), three-point percentage (3P%), free throw percentage (FT%), as well as MVP voting points won from both seasons. Our findings reveal patterns in player shooting performance, MVP points accumulation, and the intertwined effects of the transitional foul rule… We discovered significant correlations between shooting efficiency and MVP points, with distinct variations observed between the two seasons.

Understanding the implications of the transitional foul rule in comparison to the previous season is crucial for stakeholders across the NBA landscape, from players and coaches to researchers and fans. By illuminating the intricate relationship between rule changes, player performance, and individual accolades, this paper contributes to a deeper understanding of the evolving dynamics within the league.

This paper is structured into three main sections: Data, Results, and Discussion. @sec-data Data explores sources and methodology. It includes various sources used for data collection, cleaning, and analyzing processes, drawing from the dataset. In @sec-results Results, we present our findings, focusing on data visualizations and analyses. @sec-discussion Discussion provides a comprehensive analysis of our research, highlights the implications for NBA basketball, the importance of limitations and considerations in our analysis, and identifies areas for future research. @sec-conclusion Conclusion summarizes our main findings and the broader significance of our research.

@@

The data for this paper are obtained from Basketball Reference [@basketball], which provides a wide range of basketball-related data that includes statistics, player and team information, game results, and more. The datasets used here are @basketball2023 and @basketball2022. These datasets include various statistics and information about NBA players who received votes for the MVP award, as well as their performance metrics for the 2022-23 and 2021-22 seasons. The data analysis involved using the R programming language [@citeR], along with the `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `knitr` [@citeKnitr], `cowplot` [@cowplot], `broom` [@broom], and `rstanarm` [@rstanarm] packages for data collection, cleaning, analysis, and statistical modeling. Within the `tidyverse` package [@citeTidyverse], I used `readr` [@citeReadr], `dplyr` [@citeDplyr], `tidyr` [@citeTidyr], and `ggplot2` [@citeGgplot2] for reading CSV files, manipulating data, tidying data, and creating figures.

@@

## Data Visualization
### The distribution of MVP points won by top 10 players in the 2021-22 and 2022-23 seasons

@@

# Results {#sec-results}
## Summary of regression model results
Our results are summarized in @tbl-coefficients and @tbl-additional. The estimate intercept is -920.99 with a high standard error of 1833.86 and high p-value of 0.63 (significantly greater than 0.05). The negative intercept might initially suggest a baseline negative effect, but considering high standard error and p-value, the intercept is not statistically significant. This means that the intercept does not provide reliable information about the baseline level of MVP points when shooting percentages are not accounted for.

@@

## Observed versus predicted MVP points won
@fig-modelresults shows the observed MVP points (actual outcomes) against predicted MVP points (outcomes estimated by the model). The perfect predictions would result in all points falling exactly on the red dashed line, where the predicted values equal the observed values. However, as we can see, the data points are notably distant from this line, particularly at higher observed points. For instance, an observed value of 915 points in a predicted value of only 534.42. This significant deviation indicates that the model tends to underpredict MVP points for players with higher observed scores. In addition, for lower observed points, the model's performance is inconsistent. For instance, when the observed value is at 30, the predicted value is slightly negative (-5.09), which doesn't align with the possible range of MVP points. Conversely, an observed value of 3 has an inaccurate predicted value of 369.98. This means there is a potential issue of reliability and precision in the model’s predictions for top-performing players.

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

@@

Weaknesses and next steps should also be included.

# Conclusion {#sec-conclusion}

>>>>

---
title: "The Impact of the 2022-23 NBA Rule Change"
subtitle: "Exploring the Relationship Between the New Transitional Foul Rule, Player Shooting Performance, and MVP Points"
author: 
  - Amie Liu
thanks: "Code and data are available at: https://github.com/AmieLiu/the-NBA-Rule-Change.git."
date: today
date-format: long
abstract: "Rule changes in professional basketball play a critical role in shaping both strategies and player performances. Specifically, the introduction of the transitional foul rule in the 2022-23 NBA season, which changes the way fouls are called during transition plays, could significantly affect players' shooting efficiency and their prospects for MVP awards. This paper examines the implications of this rule change, comparing player statistics between the 2022-23 and 2021-22 seasons. We found no statistically significant correlations between shooting efficiency and MVP points. This paper offers insights into the interplay between rule modifications, player performance, and individual accolades, underscoring the broader significance of adapting to regulatory shifts in professional sports."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

A clear gap exists in understanding the nuanced impacts of the new transitional foul rule on player performance and MVP considerations, especially when compared to the previous season. This paper seeks to fill this gap by thoroughly analyzing the statistical data from both the 2022-23 and 2021-22 NBA seasons to highlight the changes, continuities, and emerging trends over these times. To address this gap, we used the R programming language [@citeR] to analyze shooting performance metrics, such as field goal percentage (FG%), three-point percentage (3P%), free throw percentage (FT%), as well as MVP voting points won from both seasons. Our findings reveal no statistically significant relationships between shooting efficiency and MVP points in the 2022-23 NBA season.

This paper is structured into four main sections: Data, Model, Results, and Discussion. @sec-data Data explores sources, methodology, measurement, and data visualization. It includes various sources used for data collection, cleaning, and analyzing processes, drawing from the dataset. @sec-model Model describes the statistical model employed in our analysis. In @sec-results Results, we present our findings, focusing on comprehensive analyses obtained from the model. @sec-discussion Discussion provides a summary of our research, highlights the implications for NBA basketball, the importance of limitations and considerations in our analysis, and identifies areas for future research.

@@

The data for this paper are obtained from Basketball Reference [@basketball], which provides a wide range of basketball-related data that includes statistics, player and team information, game results, and more. The datasets used here are @basketball2023 and @basketball2022. These datasets include various statistics and information about NBA players who received votes for the MVP award, as well as their performance metrics for the 2022-23 and 2021-22 seasons. The data analysis involved using the R programming language [@citeR], along with the `tidyverse` [@citeTidyverse], `janitor` [@citeJanitor], `knitr` [@citeKnitr], `cowplot` [@cowplot], `broom` [@broom], and `rstanarm` [@rstanarm] packages for data collection, cleaning, analysis, and statistical modeling. Within the `tidyverse` package [@citeTidyverse], `readr` [@citeReadr], `dplyr` [@citeDplyr], `tidyr` [@citeTidyr], and `ggplot2` [@citeGgplot2] are used for reading CSV files, manipulating data, tidying data, and creating figures.

@@

## Data Visualization

### The distribution of MVP points won by top 10 players in the 2021-22 and 2022-23 seasons

@@

# Results {#sec-results}

## Summary of regression model results

Our results are summarized in @tbl-coefficients and @tbl-additional. The estimate intercept is -920.99 with a high standard error of 1833.86 and high p-value of 0.63 (significantly greater than 0.05). The negative intercept might initially suggest a baseline negative effect, but considering high standard error and p-value, the intercept is not statistically significant. This means that the intercept does not provide reliable information about the baseline level of MVP points when shooting percentages are not accounted for.

@@

## Observed versus predicted MVP points won

@fig-modelresults shows the observed MVP points (actual outcomes) against predicted MVP points (outcomes estimated by the model). The perfect predictions would result in all points falling exactly on the red dashed line, where the predicted values equal the observed values. However, as we can see, the data points are notably distant from this line, particularly at higher observed points. For instance, an observed value of 915 points in a predicted value of only 534.42. This significant deviation indicates that the model tends to underpredict MVP points for players with higher observed scores. In addition, for lower observed points, the model's performance is inconsistent. For instance, when the observed value is at 30, the predicted value is slightly negative (-5.09), which doesn't align with the possible range of MVP points. Conversely, an observed value of 3 has an inaccurate predicted value of 369.98. This indicates that the observed MVP points and the predicted MVP points are not well aligned. Specifically, the model's predictions deviate significantly from the actual outcomes.

@@

## Summary of key findings {#sec-first-point}

The primary objective of this paper is to analyze the consequences of the 2022-23 NBA transitional foul rule on player shooting performance and MVP points won.  The methodology involves collecting and analyzing data from the NBA seasons of 2021-22 and 2022-23, sourced from @basketball. We compare player shooting performance across these two seasons to visualize the data, illustrating trends between the seasons. The key variables analyzed include FG%, 3P%, FT%, and MVP points. There's a shift in MVP points distribution, with top-ranked players in the 2022-23 season generally receiving more points compared to the previous season, suggesting potential changes in voter perception or player performance (see @fig-points).

To further examine the relationship between shooting performance metrics and MVP voting outcomes, a linear regression model is constructed. This model uses MVP points as the dependent variable and shooting percentages (FG%, 3P%, FT%) as independent variables. The use of linear regression allows for quantifying the potential impact of shooting performance on MVP voting, considering the influence of the transitional foul rule. However, the results show that there are no significant findings about relationships between shooting percentages and MVP points (see @tbl-coefficients and @tbl-additional).

## The psychological impacts of the new NBA's transitional foul rule change on players

To understand the implications of the NBA's transitional foul rule change, it is essential to consider its psychological impacts on players, akin to the effects observed with strategic timeouts. As discussed by @timeOut, strategic timeouts are often employed to disrupt a player's performance during critical moments of a game. While they are intended to disrupt a player's rhythm and impair performance, they often fail to achieve their intended effect [@timeOut]. Instead, players may adapt and focus more intensely. Similarly, with the new transitional foul rule in the NBA, players might initially experience disruptions in their game strategy but adapt quickly to optimize performance under the new conditions. This adaptability is evident in the lack of significant changes in the Three-Point Percentage (3P%) and Free Throw Percentage (FT%) among the top players (@fig-percentage).

Furthermore, the transitional foul rule could increase players' confidence during fast-break situations because they know that unnecessary contact could result in significant penalties for the opposing team. This may lead to a more fluid approach to offensive transitions, enhancing both individual and team scoring efficiency. It also allows players to play the game more freely and focus more on their performance rather than feeling defensive pressure. This trend is indicated by @fig-percentage, which shows an observable increase in the FG% of the top 10 players in the 2022-23 season compared to the previous season. The transitional foul rule in the NBA not only modifies the game's structural dynamics but also has profound psychological effects on players, influencing how players strategize and execute during games. As they adapt to these new changes, the overall impact on performance metrics and psychological resilience becomes a fascinating aspect of professional basketball dynamics.

## Potential biases in the MVP voting process

The inherent biases could affect the MVP voting process when evaluating the implications of the NBA's transitional foul rule changes. Although shooting percentages statistical metrics provide quantitative measures of player performance, there is no significant relationship between a player’s shooting performance and MVP points won, as shown by @fig-modelresults. Therefore, this means that other factors influence the allocation of MVP points, which may involve subjective judgment.

The MVP voting process is conducted by a total of 100 sportswriters and broadcasters chosen by the NBA from the United States and Canada [@mvpVoting]. Although these panelists are responsible for assessing player performances over the season [@mvpVoting], their subjective opinions and personal biases could affect their voting choices. @biases highlights similar biases in referee decisions, providing a parallel to potential biases in MVP voting. The research suggests that well-known NBA players often receive favorable calls, particularly during crucial moments in playoff games [@biases]. This observation about referee bias can be extended to the MVP voting process. As referees may show favoritism towards star players, MVP voters can also be influenced by a player's fame, media visibility, or the prevailing narrative about their season performance. This bias can result in equally deserving but less prominent players not receiving comparable recognition in the MVP voting process. The MVP voting process not only reflects statistical performance but is also shaped by the subjective interpretations and biases of the voting panel.

@@

Although our analysis aims to explore the relationships between the NBA rule change, shooting performance metrics, and MVP points, it is important to acknowledge potential weaknesses that arise. These weaknesses include the limited scope of variables considered, the size of the datasets, statistical significance, and the subjectivity in MVP voting. In our findings, we did not observe a statistically significant relationship between MVP points and a player's shooting performance. This suggests that factors beyond shooting efficiency, such as scores, defensive contributions, team success, and player injuries, also influence MVP considerations. The exclusion of these variables may limit the depth and comprehensiveness of the analysis. In addition, our analysis is based on data from the 2022-23 and 2021-22 NBA seasons, which may not provide a sufficiently large sample size to draw definitive conclusions. Therefore, expanding the time frame to include data from a longer period, such as 10 years, could enhance the robustness of our key findings. This can allow us to comprehensively understand the impact of rule changes on player performance and MVP points. Moreover, the MVP voting process is subjective and may be influenced by the personal biases of the voting panel. This subjectivity leads to uncertainty in the interpretation of MVP points outcomes.

To deepen our understanding of the factors influencing MVP points and the effects of rule changes in the NBA, future research is needed. Firstly, a broader range of player performance metrics beyond shooting percentages is necessary to grasp the player's overall impact on the game. Secondly, understanding the correlation between team success and individual player performance in MVP considerations is essential, as MVP voting considers a player's contribution to their team's success. Thirdly, extending the time frame of datasets to identify long-term trends in MVP voting and player performance is necessary to track the evolution of player recognition over time. Finally, exploring the composition and dynamics of the MVP voting panel, including individual biases, preferences, and voting patterns, will facilitate a deeper understanding of their influence on MVP outcomes. This investigation can encompass demographic factors, geographic biases, and media narratives that may impact voting decisions.
